Static contract checking for Haskell

### 2011-12-22 paper/paper.tex
Paper: 'min' and lemmas.
A little more on 'min', and a description of lemmas. The
lemma description is short, and Koen has a much more complete
description in the google doc.

We introduce a predicate $\min$, short for ``am interested'', and use
it both to guide proof search and guarantee (?) the existence of
finite models.  Dimitrios says that such use of a predicate, at least
the guiding of proof search part, is related to the concept of
``triggers''.  We are particularly interested in finite
\emph{counter}-models, which exist when the contracts aren't provable
(i.e. when our refutation theory is counter satisfiable).

@@

The $\min$ predicate guarantees finite models by preventing us from
proving all models are infinite (recall the completeness theorem for
FOL).  Without $\min$, the axiomitization of any infinite datatype $T$
would allow us to prove $T$, and hence the domain of our models, was
infinite.  Namely, we axiomitize the injectivity and disjointness of
$T$'s term constructors, and without $\min$ these axioms prove that
inductive term constructors are injective, but not surjective, and
hence the domain is infinite.  However, we guard these axioms, namely
$\phiInjective$ and $\phiDisjoint$ in Figure~\ref{fig:dtrans}, with
$\min$ predicates, to restrict their use to interesting constructor
applications.

@@

\item $(\phiLazy)$ No full application of a constructor
  $K_i$ is unreachable or crashing.
\end{itemize}

@@

\subsection{$\ctrans{}$ -- Contracts}\label{sec:contracts}
\note[nc]{%
Figure~\ref{fig:newCTrans} contains the $\min$-translation of contracts, and
Figure~\ref{fig:naiveCTrans} contains the naive contract translation,
with no $\min$'s.  The min translation reduces to the naive translation when
all $\min$'s are assumed true, i.e. when an axiom $\forall x.\min{x}$ is added.

The min translation has two forms, $\ctrans{}^-$ for
assumptions/axioms and $\ctrans{}^+$
for conclusions/theorems.  The choice of minus and 
plus comes from the negative and positive
positions in an implication: $- \to +$.  The goal of the $\min$'s twofold: 
(1) to allow finite counter models when the theory 
does not prove the contracts; (2) to limit the
applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ttrans{}$), 
in order to prune the proof search space.  The $\min$'s
appear as obligations (the premises in implications) 
in contract and definition axioms. The mnemonic is ``(a)m in(terested)'',
and the translation is designed so that $\min{e}$ is introduced whenever
$e$ is evaluated or inspected.  The axiom translations contain $\min{}$
obligations that prevent using the axioms on ``uninteresting'' terms.

@@

Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
Figure~\ref{fig:newCTrans}.  When proving that $e \in \{x|p\}$, we get to
assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
we need to prove we're interested in $e$.  The $\min{e}$ constraint
here is not optional, because in addition to concluding 
$e \neq \unr \to p[e/x] \in \{\true,\unr\}$, we also conclude $\min{p[e/x]}$.
The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
$p[e/x] \in \{\true,\unr\}$, and because it can be used to unfold definitions,
we guard it with $\min{e}$.
}

@@

Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
it's natural to think of contract assumptions in the theory as
negative, and contract goals as positive.  An example is helpful:
\protect \begin{eqnarray*}
\ctrans{g \in \cfc \to \cfc}^+ &=& \forall x. \cf{x} \to (\min{g ~x} \to \cf{g ~x}),\\
&&\text{whereas}\\
\ctrans{g \in \cfc \to \cfc}^- &=& \forall x. \min{g ~x} \to (\min{x} \to \cf{x}) \to \cf{g ~x}.
\end{eqnarray*}
Now, given $\ctrans{g \in \cfc \to \cfc}^-$, we may 
want to prove $\ctrans{g ~(g ~Z) \in \cfc}^+$.
So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
$\min{g ~Z} \to \cf{g ~Z}$
But that follows by instantiating $g$'s contract again, and 
then using $\cf{Z}$ to get $\min{Z} \to \cf{Z}$.

@@

  the theory $T$ should consist of $\in^-$ axioms.
  We want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
  But
  that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.

@@

@@

\begin{figure*}
$$ \forall x_1,\dots,x_n.~\full{f}(x_1,\dots,x_n) = f~x_1~x_2~\dots~x_n$$
$$ \cf{f} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{\full{f}(x_1,\dots,x_n)}$$
$$\forall g,x.~\cf{f} \land \cf{x} \to \cf{g~x}$$
\caption{Encoding of higher-orderness}
\label{ho-fig}
\end{figure*}

>>>>

We introduce a predicate $\min$, short for ``am interested'', and use
it both to guide proof search and guarantee (?)\footnote{
  % 
  Do we really guarantee finite models?  I think Koen said ``yes'',
  but I don't see how to prove this.  Namely, I don't see how to prove
  we never have $S : M \to M$ for $M \subseteq \{ x \mid \min{x} \}$ and
  $M \neq \emptyset$.

  I think I wrote down an example where we have $\min{Z}$,
  $\min{S~Z}$, $\min{S~(S~Z)}$, etc, but it seems I'd need an
  induction axiom to prove that $S$ really took $M$ to $M$, for $M$
  the naturals.
  % 
} the existence of
finite models.  Dimitrios says that such use of a predicate, at least
the guiding of proof search part, is related to the concept of
``triggers''.  We are particularly interested in finite
\emph{counter}-models, which exist when the contracts aren't provable
(i.e. when our refutation theory is counter satisfiable).

@@

The $\min$ predicate guarantees (?) finite models by preventing us from
proving all models are infinite (recall the completeness theorem for
FOL).  Without $\min$, the axiomitization of any infinite datatype $T$
would allow us to prove $T$, and hence the domain of our models, was
infinite.  Namely, we axiomitize the injectivity and disjointness of
$T$'s term constructors, and without $\min$ these axioms prove that
inductive term constructors are injective, but not surjective, and
hence the domain is infinite.  However, we guard these axioms, namely
$\phiInjective$ and $\phiDisjoint$ in Figure~\ref{fig:dtrans}, with
$\min$ predicates, to restrict their use to interesting constructor
applications.

The $\min$ predicate complicates the translation of contracts,
requiring us to define separate contract translations for goals and
axioms.  The only actual difference is for predicate-contract, but we
have considered many minor variations.

@@

\item $(\phiLazy)$ No full application of a constructor $K_i$ is
  unreachable or crashing.  In other words, term constructors don't
  scrutinize their arguments / are lazy.
\end{itemize}

@@

\subsection{$\ctrans{}$ -- Contracts}\label{sec:contracts}
Figure~\ref{fig:newCTrans} contains the $\min$-translation of contracts, and
Figure~\ref{fig:naiveCTrans} contains the naive contract translation,
with no $\min$'s.  The min translation reduces to the naive translation when
all $\min$'s are assumed true, i.e. when an axiom $\forall x.\min{x}$ is added.

The min translation has two forms, $\ctrans{}^-$ for
assumptions/axioms and $\ctrans{}^+$
for conclusions/theorems.  The choice of minus and 
plus comes from the negative and positive
positions in an implication
\[
- \to +.
\]
The goal of the $\min$'s twofold: 
(1) to allow finite counter models when the theory 
does not prove the contracts; (2) to limit the
applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ttrans{}$), 
in order to prune the proof search space.  The $\min$'s
appear as obligations (the premises in implications) 
in contract and definition axioms. The mnemonic is ``(a)m in(terested)'',
and the translation is designed so that $\min{e}$ is introduced whenever
$e$ is evaluated or inspected.  The axiom translations contain $\min{}$
obligations that prevent using the axioms on ``uninteresting'' terms.

@@

Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
Figure~\ref{fig:newCTrans}.  When proving that $e \in \{x|p\}$, we get to
assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
we need to prove we're interested in $e$.  The $\min{e}$ constraint
here is not optional, because in addition to concluding 
$e \neq \unr \to p[e/x] \in \{\true,\unr\}$, we also conclude $\min{p[e/x]}$.
The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
$p[e/x] \in \{\true,\unr\}$, and because it can be used to unfold definitions,
we guard it with $\min{e}$.

@@

Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
it's natural to think of contract assumptions in the theory as
negative, and contract goals as positive. So, the theory
consists of $\ctrans{}^-$ axioms, so call it $\Gamma^{\in^-}$.
We want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
But that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$,
and that is the refutation we feed to the theorem prover.

An example: 
given $\ctrans{g \in \cfc \to \cfc}^-$, we may 
want to prove $\ctrans{g ~(g ~Z) \in \cfc}^+$.
So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
$\min{g ~Z} \to \cf{g ~Z}$.
But that follows by instantiating $g$'s contract again, and 
then using $\phiCF$ for $Z$, which is $\ctrans{Z \in \cfc}^-$, 
i.e. $\min{Z} \to \cf{Z}$, exactly what we need.

@@

@@

\section{Implementation}

\subsection{Case translations}
XXX: describe the various case translations here.

\section{Lemmas}

It's possible to encode, prove, and apply lemmas in our system,
without adding any extensions.  The application of lemmas can change
the termination behavior of programs (although so far I think we've
only used lemmas in other lemmas), and it's up to the user to look out
for this.

Before presenting the solution we have to understand the problem: the
FOL prover requires finite axiomitizations, and hence there is no
general induction support (FOL does not allow quantification over
formulas, and so induction can only be axiomitized with a schema).
Each inductive proof of a universal property $\forall x. \phi(x)$
requires its own induction axiom, and the system does not attempt to
discover these (a separate problem).  When checking the contract of a
recursive function, we axiomitize induction for that function,
following the structure of its recursion.  But a contract may depend
on a lemma, itself requiring an inductive proof, and so we need a
facility for introduces auxiliary inductions.  Of course, we also need
to be able to state the lemmas themselves, because the system does not
discover them.

Our solution is to define ``dummy'' recursive functions, whose
recursion structure follows the induction needed to prove the lemma.
We don't actually care what these functions compute, so we define a
unit type, $QED$, and return that in the base cases.  E.g., if we want
to prove that $Z$ is a right identity for addition, then we can define
a dummy function \texttt{lem\_add\_Z\_r}
\begin{verbatim}
lem_add_Z_r x = case x of
  Zero -> QED
  Succ x_ -> lem_add_Z_r x_

\end{verbatim}
and give it a contract stating the property in question
\begin{verbatim}
lem_add_Z_r ::: x:CF -> CF&&{qed: x `eqNat` (x `add` Zero)}
\end{verbatim}

To use the lemma we, need to have $\min$ on it's instantiation.  To
get this we introduce another dummy function, $using$,
s.t. $e `using` l$ scrutinizes $l$ and returns $e$.  I.e.
\begin{verbatim}
e `using` lem = case lem of
  QED -> e
\end{verbatim}
The \texttt{case} match in $using$ introduces $\min$ for the lemma,
allowing it's contracts to be unfolded.

XXX: Koen has a long and detailed writeup of lemmas in the his google
doc.

@@

\begin{figure*}
$$ \forall x_1,\dots,x_n.~\full{f}(x_1,\dots,x_n) = f~x_1~x_2~\dots~x_n$$
$$ \cf{f} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{\full{f}(x_1,\dots,x_n)}$$
$$\forall g,x.~\cf{g} \land \cf{x} \to \cf{g~x}$$
\caption{Encoding of higher-orderness.
%
XXX: need 'min's.
%
XXX: i don't think the second and third formulas are implemented.}
\label{ho-fig}
\end{figure*}
